---
title: "IRIS Clusters Analysis Dashboard"
output: 
  flexdashboard::flex_dashboard:
    navbar:
      - { href: "https://www.linkedin.com/in/seif-kungulio/", 
      icon: "fa-linkedin" }
      - { href: https://github.com/shkungulio, 
      icon: "fa-github" }
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
runtime: shiny
---

```{r setup, include=FALSE}
# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

#Install the following libraries if required
# List of required packages
required_pkgs <- c("flexdashboard", "shiny", "tidyverse", "cluster", "factoextra", "fpc", "DT", "plotly")

# Install and load each package
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Global options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))
```

```{r}
# Data & model objects
data(iris)

# Select numeric features only
iris_features <- iris %>%
select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)

# Apply standardization
iris_scaled <- scale(iris_features)

# K-Means clustering with k = 3
set.seed(123)  # For reproducibility
kmeans_model <- kmeans(iris_scaled, centers = 3, nstart = 25)

# Attach clusters back to original data
iris_kmeans <- iris %>%
  mutate(KMeansCluster = factor(kmeans_model$cluster),
         Species = factor(Species))

# PCA for visualization
pca_model <- prcomp(iris_scaled, center = TRUE, scale. = TRUE)
pca_scores <- as_tibble(pca_model$x[, 1:2]) %>%
  rename(PC1 = PC1, PC2 = PC2)

# Analysis ready table
iris_aug <- bind_cols(iris_kmeans, pca_scores)

# Cluster centers for reference
centers_tbl <- as_tibble(kmeans_model$centers) %>%
  mutate(Cluster = factor(row_number())) %>%
  relocate(Cluster)

# Adjusted Rand Index helper
compute_ari <- function(cluster_labels, dist_matrix, true_labels) {
  cluster.stats(d = dist_matrix,
                clustering = cluster_labels,
                alt.clustering = as.numeric(factor(true_labels))
  )$corrected.rand
}
```

```{r}
# Shared reactive filters for the whole dashboard
filtered_data <- reactive({
  df <- iris_aug
  
  # Filter by species (if user selected any)
  if (!is.null(input$species_filter) && length(input$species_filter) > 0) {
    df <- df %>% filter(Species %in% input$species_filter)
  }
  
  # Filter by cluster (if user selected any)
  if (!is.null(input$cluster_filter) && length(input$cluster_filter) > 0) {
    df <- df %>% filter(KMeansCluster %in% input$cluster_filter)
  }
  
  df
  }
)

# Distance & diagnostics for filtered subset
filtered_dist <- reactive({
  df <- filtered_data()
  if (nrow(df) < 3) {
    return(NULL)
    x <- df %>%
      select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) %>%
      scale()
    dist(x)
  }
})

avg_sil_width <- reactive({
  d <- filtered_dist()
  df <- filtered_data()

if(!is.null(d) || length(unique(df$KMeansCluster)) > 2 ||
   length(unique(df$Species)) < 2) return(NA_real_)
compute_ari(df$KMeansCluster, d, df$Species)
})

color_vector <- reactive({
  df <- filtered_data()
  if (is.null(input$color_by) || input$color_by == "cluster") df$KMeansCluster else df$Species
})

legend_title <- reactive({
if (is.null(input$color_by) || input$color_by == "cluster") "Cluster" else "Species"
})
```


Sidebar {.sidebar}
-------------------------------------
### Controls
```{r sidebar_controls}

# 
radioButtons(
inputId = "color_by",
label = "Color points by",
choices = c("K-Means Cluster" = "cluster", "Species" = "species"),
selected = "cluster"
)

# 
checkboxGroupInput(
inputId = "species_filter",
label = "Species",
choices = levels(iris$Species),
selected = levels(iris$Species),
inline = FALSE
)

#
checkboxGroupInput(
inputId = "cluster_filter",
label = "Clusters (k = 3)",
choices = levels(iris_kmeans$KMeansCluster),
selected = levels(iris_kmeans$KMeansCluster),
inline = FALSE
)
```



Row {data-height=120}
-------------------------------------
### Dataset
```{r}
renderValueBox({
df <- filtered_data()
valueBox(
value = nrow(df),
caption = "Observations in view",
icon = "fa-database"
)
})
```

### Species
```{r}
renderValueBox({
df <- filtered_data()
valueBox(
value = length(unique(df$Species)),
caption = "Species present",
icon = "fa-leaf"
)
})
```

### Clusters
```{r}
renderValueBox({
df <- filtered_data()
valueBox(
value = length(unique(df$KMeansCluster)),
caption = "Clusters present",
icon = "fa-circle-nodes"
)
})
```

### Silhouette
```{r}
renderValueBox({
s <- avg_sil_width()
valueBox(
value = ifelse(is.na(s), "—", sprintf("%.3f", s)),
caption = "Avg silhouette width",
icon = "fa-chart-line"
)
})
```

Row {data-height=360}

Column {.tabset}
-------------------------------------

### **K-Means View (PCA Space)**
```{r}
renderPlotly({
df <- filtered_data()
req(nrow(df) > 1)

p <- ggplot(df, aes(x = PC1, y = PC2, color = color_vector())) +
geom_point(alpha = 0.85, size = 2.6) +
stat_ellipse(aes(group = KMeansCluster), linetype = 2, linewidth = 0.6, show.legend = FALSE) +
labs(
title = "Filtered Data in PCA Space",
subtitle = "Ellipses are drawn by K-Means clusters (k = 3)",
x = "PC1",
y = "PC2",
color = legend_title()
) +
theme(legend.position = "bottom")

ggplotly(p, tooltip = c("x", "y")) %>% layout(legend = list(orientation = "h"))
})
```
> K-Means clustering visualized in PCA space with ellipses representing clusters.

### **Silhouette Plot**
```{r}
renderPlot({
d <- filtered_dist()
df <- filtered_data()

validate(
need(!is.null(d), "Select at least 3 observations to compute silhouette."),
need(length(unique(df$KMeansCluster)) >= 2, "Silhouette needs at least 2 clusters in the filtered view.")
)

sil <- cluster::silhouette(as.numeric(df$KMeansCluster), d)
factoextra::fviz_silhouette(sil) +
labs(title = "Silhouette Widths (Filtered View)")
})
```
> Silhouette plot showing clustering quality for the filtered dataset.

### ARI vs. Species
```{r}
renderPlot({
a <- ari_kmeans()
validate(need(!is.na(a), "ARI needs at least 2 species and 2 clusters in the filtered view."))

tibble(metric = "Adjusted Rand Index", value = a) %>%
ggplot(aes(x = metric, y = value)) +
geom_col() +
coord_flip() +
scale_y_continuous(limits = c(0, 1)) +
labs(title = "Agreement Between K-Means and True Species (Filtered)")
})
```


Column {.tabset, data-height=280}
-------------------------------------

### **K-Means Cluster Plot**
```{r}
renderPlotly({
  # K-Means cluster visualization using factoextra
  p <- fviz_cluster(
    kmeans_model,
    data = iris_scaled,
    geom = "point",
    ellipse.type = "norm",
    main = "K-Means Clustering on Standardized Iris Features"
)
# Convert to plotly for interactivity
ggplotly(p)
})
```
> K-Means clustering results visualized with cluster ellipses.

### **Silhouette Plot (k = 3)**
```{r}
renderPlot({
# Base silhouette plot
  suppressWarnings(
    fviz_silhouette(
      sil_obj,
      main = "Silhouette Plot for K-Means (k = 3)")
  )
})
```
> Silhouette plot showing the quality of clustering for k = 3.


Column {.tabset}
-------------------------------------

### **Cluster Centers (Standardized Scale)**
```{r}
renderDataTable({
datatable(
centers_tbl,
rownames = FALSE,
options = list(pageLength = 5)
)
})
```
> K-Means cluster centers on the standardized feature scale.

### **Iris Data with Cluster Assignment**
```{r}
renderDataTable({
df <- filtered_data()
datatable(
df,
options = list(pageLength = 10)
)
})
```
> Original iris dataset with K-Means cluster assignments, filtered by user selections.
