---
title: "IRIS Clusters Analysis Dashboard"
output:
  flexdashboard::flex_dashboard:
    navbar:
    - href: "https://www.linkedin.com/in/seif-kungulio/"
      icon: "fa-linkedin"
    - href: https://github.com/shkungulio
      icon: "fa-github"
    orientation: rows
    vertical_layout: scroll
    theme: united
runtime: shiny
resource_files:
- ilogo.png
---

```{r setup, include=FALSE}
#Install the following libraries if required
library(shiny)
library(flexdashboard)
library(tidyverse)
library(cluster)
library(factoextra)
library(fpc)
library(DT)
library(plotly)

options(scipen = 999)
theme_set(theme_minimal(base_size = 12))



# Data & model objects
data(iris)

# Select numeric features only
iris_features <- iris %>%
  select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)

# Load saved scaling parameters and trained K-Means model
scale_center <- readRDS("../resources/scale_center.rds")
scale_scale  <- readRDS("../resources/scale_scale.rds")
kmeans_model <- readRDS("../resources/kmeans_model.rds")

# Apply standardization
iris_scaled <- scale(iris_features, 
                     center = scale_center, 
                     scale = scale_scale)

# Attach clusters back to original data
iris_kmeans <- iris %>%
  mutate(KMeansCluster = factor(kmeans_model$cluster),
         Species = Species
         )

# Distance matrix for cluster diagnostics
dist_iris <- dist(iris_scaled)

# Silhouette for k = 3
sil_obj <- cluster::silhouette(kmeans_model$cluster, dist_iris)
avg_sil_width <- mean(sil_obj[, "sil_width"])

# Adjusted Rand Index helper
compute_ari <- function(cluster_labels, dist_matrix, true_labels) {
  cluster.stats(d = dist_matrix, clustering = cluster_labels,
                alt.clustering = as.numeric(true_labels)
  )$corrected.rand
  }

ari_kmeans <- compute_ari(kmeans_model$cluster, dist_iris, iris$Species)

# PCA for visualization (same approach as in your report)
pca_model <- prcomp(iris_scaled, center = TRUE, scale. = TRUE)

pca_data <- as_tibble(pca_model$x[, 1:2]) %>%
  mutate(Cluster = factor(kmeans_model$cluster), Species = iris$Species)

# Cluster centers as tibble (for display)
centers_tbl <- as_tibble(kmeans_model$centers) %>%
  mutate(Cluster = factor(row_number())) %>%
  relocate(Cluster)
```

```{r}
# Shared reactive filters for the whole dashboard
filtered_data <- reactive({
  df <- iris_kmeans

# Filter by species (if user selected any)
  if (!is.null(input$species_filter) && length(input$species_filter) > 0) {
    df <- df %>% filter(Species %in% input$species_filter)
}

# Filter by cluster (if user selected any)
  if (!is.null(input$cluster_filter) && length(input$cluster_filter) > 0) {
    df <- df %>% filter(KMeansCluster %in% input$cluster_filter)
}

  df
})

filtered_pca <- reactive({
  df <- pca_data
  
  if (!is.null(input$species_filter) && length(input$species_filter) > 0) {
    df <- df %>% filter(Species %in% input$species_filter)}

  if (!is.null(input$cluster_filter) && length(input$cluster_filter) > 0) {
    df <- df %>% filter(Cluster %in% input$cluster_filter)}

  df
})
```


Sidebar {.sidebar data-width=240}
--------------------------------------------------------------------------------
### **Controls**

```{r sidebar-controls}
radioButtons(
  inputId = "color_by",
  label   = "Color points by",
  choices = c("K-Means Cluster" = "cluster", "Species" = "species"),
  selected = "cluster"
)

checkboxGroupInput(
  inputId = "species_filter",
  label   = "Filter by Species",
  choices = levels(iris$Species),
  selected = levels(iris$Species)
)

checkboxGroupInput(
  inputId = "cluster_filter",
  label   = "Filter by Cluster",
  choices = levels(iris_kmeans$KMeansCluster),
  selected = levels(iris_kmeans$KMeansCluster)
)

sliderInput(
inputId = "point_size",
label = "Point size",
min = 1,
max = 6,
value = 2.5,
step = 0.5
)
```



***



&nbsp; © [2025 Seif Kungulio](http://shkungulio.com)  


Row {data-height=110}
--------------------------------------------------------------------------------
### Tobal Observations
```{r}
valueBox(
value   = nrow(iris),
caption = "Total Observations",
icon    = "fa-database"
)
```

### True Species
```{r}
valueBox(
value   = length(unique(iris$Species)),
caption = "True Species",
icon    = "fa-leaf"
)
```

### K-Means Clusters
```{r}
valueBox(
value   = length(unique(kmeans_model$cluster)),
caption = "K-Means Clusters (k)",
icon    = "fa-object-group"
)
```

### Adjusted Rand Index
```{r}
valueBox(
value   = sprintf("%.3f", ari_kmeans),
caption = "Adjusted Rand Index",
icon    = "fa-check-circle"
)
```

### Avg. Silhouette Width
```{r}
valueBox(
value   = sprintf("%.3f", avg_sil_width),
caption = "Avg. Silhouette Width",
icon    = "fa-chart-line"
)
```


Row {data-height=320}
--------------------------------------------------------------------------------
### **Cluster Scatterplot (Petal measurements)**
```{r}
renderPlotly({
  df <- filtered_data()
  req(nrow(df) > 1)

  p <- ggplot(
    df,
    aes(
      x = Petal.Length,
      y = Petal.Width,
      color = color_vector(),
      text = paste(
        "Sepal.Length:", Sepal.Length,
        "<br>Sepal.Width:", Sepal.Width,
        "<br>Petal.Length:", Petal.Length,
        "<br>Petal.Width:", Petal.Width,
        "<br>Species:", Species,
        "<br>Cluster:", KMeansCluster
      )
    )
  ) +
    geom_point(alpha = 0.85, size = 2.8) +
    labs(
      title = "Petal Length vs Petal Width (Filtered)",
      x = "Petal.Length",
      y = "Petal.Width",
      color = legend_title()
    ) +
    theme(legend.position = "right")

  ggplotly(p, tooltip = "text") %>% layout(legend = list(orientation = "h"))
})
```
> Scatterplot of Petal Length vs Petal Width for the filtered iris data, colored by either K-Means clusters or true species labels.

### **K-Means View (PCA space)**
```{r}
renderPlotly({
  df <- filtered_data()
  req(nrow(df) > 1)

  p <- ggplot(df, aes(x = PC1, y = PC2, color = color_vector())) +
    geom_point(alpha = 0.85, size = 2.6) +
    stat_ellipse(aes(group = KMeansCluster), linetype = 2, linewidth = 0.6, show.legend = FALSE) +
    labs(
      title = "Filtered Data in PCA Space",
      subtitle = "Ellipses are drawn by K-Means clusters (k = 3)",
      x = "PC1",
      y = "PC2",
      color = legend_title()
    ) +
    theme(legend.position = "bottom")

  ggplotly(p, tooltip = c("x", "y")) %>% layout(legend = list(orientation = "h"))
})
```
> Scatterplots showing the filtered iris data colored by either K-Means clusters or true species labels.



Row {data-height=320}
---
### Chart 3
```{r}

```
>

### Chart 4
```{r}
renderPlot({

# Base silhouette plot still useful for overall structure

plot(
sil_obj,
main = "Silhouette Plot for K-Means (k = 3)",
border = NA
)
})
```
> 

Row {.tabset .tabset-fade data-height=240}
---
### Table 1
```{r}

```

### Table 2
```{r}

```